---
title: "Reproducible Research: Course Project 1"
output: html_document
---

##Assignment
This assignment will be described in multiple parts. You will need to write a report that answers the questions detailed below. Ultimately, you will need to complete the entire assignment in a single R markdown document that can be processed by knitr and be transformed into an HTML file.

The submission must include the following

1. Code for reading in the dataset and/or processing the data
2. Histogram of the total number of steps taken each day
3. Mean and median number of steps taken each day
4. Time series plot of the average number of steps taken
5. The 5-minute interval that, on average, contains the maximum number of steps
6. Code to describe and show a strategy for imputing missing data
7. Histogram of the total number of steps taken each day after missing values are imputed
8. Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends
9. All of the R code needed to reproduce the results (numbers, plots, etc.) in the report

###Read the data
```{r}
if (!file.exists('activity.csv')) {
  unzip(zipfile = "repdataFdataFactivity.zip")
}

activity_info <- read.csv("activity.csv", header = T)
```

###Total number of steps per day
```{r}
# Calculate the total steps taken per day
total_Steps <- aggregate(steps ~ date, activity_info, FUN=sum, na.rm=T)
head(total_Steps)
library(ggplot2)
ggplot(total_Steps, aes(x=steps)) + geom_histogram(bins=10, fill="lightseagreen", colour= "darkslategray4") +
      xlab("Total Steps") + ylab("Frequency") + ggtitle("Histogram of Total steps per day")

```

###Mean and median number of steps taken each day
For the mean and median we will use the summary command. The mean and median can be read from the summary, this is the mean and median excluding the NAs.
```{r}
with(total_Steps, { stats <- summary(steps)
stats })

# You can also calculate the mean and median directly as follows
# mean_Steps <- mean(total_Steps$steps, na.rm=T)
# mean_Steps
# median_Steps <- median(total_Steps$steps, na.rm=T)
# median_Steps
```

###What is the average daily activity pattern?
```{r}
# Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps
# taken, averaged across all days (y-axis)
average_daily_activity <- aggregate(steps ~ interval, activity_info, FUN=mean, na.rm=T)
tail(average_daily_activity)
ggplot(average_daily_activity, aes(x=interval, y=steps)) + geom_line(colour="cornflowerblue") + 
      ggtitle("Average Daily Activity Pattern") + xlab("5-minute interval") +
      ylab("Average number of steps taken")

#Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
max_interval <- average_daily_activity[which.max(average_daily_activity$steps),]
max_interval$interval
```

###Imputing missing values
```{r}
#The total number of rows with NAs - if you look at the data only the steps field has NAs
colSums(is.na(activity_info))
num_NAs <- sum(is.na(activity_info))
num_NAs

no_NAs_activity <- activity_info
new_val <- mean(activity_info$steps, na.rm = T)
# na_place <- which(is.na(activity_info))
# length(na_place)
# vec_of_means <- rep(new_val, times=length(na_place))
# no_NAs_activity[na_place, "steps"] <- vec_of_means


no_NAs_activity$steps[is.na(no_NAs_activity$steps)] <- new_val
no_NAs_activity_tot <- aggregate(steps ~ date, no_NAs_activity, FUN=sum, na.rm=T)
ggplot(no_NAs_activity_tot, aes(x=steps)) + geom_histogram(bins=10, fill="lightseagreen", 
      colour= "darkslategray4") +
      xlab("Total Steps") + ylab("Frequency") + 
      ggtitle("Histogram of Total steps per day (NAs replaced with mean)")

#The mean and median are still similar to before
with(no_NAs_activity_tot, { stats <- summary(steps)
stats })
```

###Are there differences in activity patterns between weekdays and weekends?
```{r}
library(dplyr)
no_NAs_activity$day <- weekdays(as.Date(no_NAs_activity$date))
#head(no_NAs_activity)
no_NAs_activity$dayType <- with(no_NAs_activity, day=="Saturday" | day=="Sunday")
head(no_NAs_activity)
no_NAs_activity <- mutate(no_NAs_activity,dayType = ifelse(dayType==T, "weekend", "weekday"))
head(no_NAs_activity)
no_NAs_activity_mean <- aggregate(no_NAs_activity$steps, 
                      list(interval = as.numeric(as.character(no_NAs_activity$interval)), 
                           dayType = no_NAs_activity$dayType),
                      FUN = "mean")
head(no_NAs_activity_mean)
names(no_NAs_activity_mean)[3] <- "mean_steps"
library(lattice)
xyplot(mean_steps ~ interval | dayType, data = no_NAs_activity_mean, type = "l", layout = c(1,2))
```
